Evolutionary rescue and reintroduction of resistant frogs allows recovery in the presence of a lethal fungal disease

### 2023-11-21 doc/manuscript/translocation.qmd
Describe objectives and edit translocation results

<!--# I've reworded this paragraph somewhat to reflect the revised order of sections and to better highlight the key outcomes of recovery, persistence, and evolution. Consider rewording to provide explicit objectives of each of the sections (plus summary of results). E.g., for genomics section objective was to provide insights into the evolutionary response by MYL frogs to Bd infection across their range in the Sierra Nevada, and thereby provide a mechanism by which frogs reintroduced into sites in Yosemite from which they were extirpated by Bd could recover (instead of being driving to extirpation again).-->

In the current study, we combine results from a 15-year landscape-scale frog reintroduction effort, a detailed population model, and genomic analyses to provide an unprecedented example of how reintroduction of resistant/tolerant frogs obtained from recovering populations [@knapp2016] allows the successful reestablishment of extirpated MYL frog populations despite ongoing disease.
Specifically, the reintroduction of resistant/tolerant frogs to 24 sites and results from long-term capture-mark-recapture (CMR) surveys indicate the capacity of reintroduced populations to become established and eventually recover.
We used these data to parameterize a population model that extends our inferences well beyond the temporal extent of the CMR data, and shows that the recovering populations are likely to persist over a 50-year period.
Finally, genomic results obtained using exome capture methods highlight evolutionary rescue as potentially key mechanism by which MYL frogs are able to recover in the presence of ongoing Bd infection.
These results show that frogs have undergone substantial evolutionary change associated with Bd exposure, changes that might provide frogs with increased resistance/tolerance to disease.
Collectively, these results provide one of the first examples of how evolutionary rescue might facilitate the successful reintroduction of resistant/tolerant animals and allow the large-scale recovery of disease-impacted species.

@@

@@

Of the `r nrow(d2)` translocated populations, `r frog_counts_gt0_nrow` (`r frog_counts_gt0_nrow / nrow(d2)`) showed evidence of successful reproduction in subsequent years, as indicated by the presence of tadpoles and/or juveniles.
For these `r frog_counts_gt0_nrow` populations, one or both life stages were detected in nearly all survey-years following translocation (proportion of survey-years: median = `r frog_counts_gt0_median`, range = `r frog_counts_gt0_min`--`r frog_counts_gt0_max`).
These same populations were also those in which recruitment of new adults (i.e., progeny of translocated individuals) was detected.
As with early life stages, recruits were detected in the majority of post-translocation survey-years (proportion of survey-years: median = `r frog_captures_gt0_median`, range = `r frog_captures_gt0_min`--`r frog_captures_gt0_max`).
In summary, survey results indicate that translocations resulted in the establishment of reproducing MYL frog populations at most recipient sites despite the ongoing presence of Bd.

@@

Results from frog translocations indicated that adult survival was often relatively high and that most populations showed evidence of successful reproduction and recruitment (described above).
Although suggestive of population establishment, a decade or more of surveys may be necessary to confirm that populations are in fact self-sustaining [@joseph2018].
To extend our inferences of population establishment beyond those possible from the 2-16 years of site-specific CMR data, we developed a population viability model.
Specifically, to test whether the observed yearly adult survival probabilities in translocated populations were sufficient to promote long-term viability, we built a stage-structured matrix model that captured known frog demography and included demographic and environmental stochasticity.
We parameterized the model using CMR data from translocated populations and known life history values in this system (@tbl-param_values).

@@

Reintroduction of resistant/tolerant MYL frogs was remarkably successful in reestablishing viable populations in the presence of Bd.
Of the 12 translocated populations, approximately 80% showed evidence of both successful reproduction and recruitment of new adults.
Bd loads measured before and after translocation were similar and generally remained well below the level at which frog mortality occurs.
Annual survival for 12 of the 24 translocated cohorts exceeded 50%, and \> 70% of translocated cohorts had survival above this 50% level when the earliest translocations are excluded (i.e., translocations conducted when methods were still being refined).
The fact that the relatively low Bd loads and correspondingly high frog survival in the presence of Bd was maintained between the natal/donor sites and the recipient sites indicates that these characteristics of naturally-recovering populations were not solely an effect of the natal/donor site or site \* frog interactions, but instead were due at least in part to resistance/tolerance inherent in the frogs.
In addition, the relatively high survival of translocated frogs was maintained in their progeny, as expected if resistance/tolerance has a genetic basis.
Results from the population viability model were also encouraging.
In particular, translocated populations with \> 50% survival in the first year post-translocation were predicted to have a low probability of extinction over 50 years (probability of extinction = 0 - 0.32).
In light of the generally low success rate of amphibian reintroduction efforts [@dodd2005], our success in reestablishing MYL frog populations via translocation of resistant/tolerant individuals is striking, and even more so given that this success occurred despite ongoing infection of frogs with Bd.

>>>>

<!--# I've reworded the next paragraph to provide explicit objectives of each of the sections (plus summary of results). -->

In the current study, we had three primary objectives.
First, to determine whether the reintroduction of resistant/tolerant MYL frogs obtained from populations recovering from Bd-caused declines [@knapp2016] allows the successful reestablishment of extirpated MYL frog populations despite ongoing disease, we conducted a 15-year landscape-scale frog reintroduction effort.
Second, to extend our inferences of population recovery well beyond the temporal extent of our multi-year survey data, we developed a model to estimate the probability of persistence for the reintroduced populations over a multi-decadal period.
Third, we used exome capture methods to provide insights into the evolutionary response by MYL frogs to Bd infection, and thereby provide a mechanism by which frogs reintroduced into sites from which they were previously extirpated by Bd could recover (instead of being driving to extirpation again).
Our results indicate the capacity of reintroduced populations to become established and eventually recover despite ongoing disease, that the recovering populations are likely to persist over a 50-year period, and that frogs have undergone substantial evolutionary change associated with Bd exposure, changes that might provide frogs with increased resistance/tolerance to disease.

<!--# Previous final paragraph of Introduction (revised post-review to reflect reordered sections): In the current study, we combine results from a 15-year landscape-scale frog reintroduction effort, a detailed population model, and genomic analyses to provide an unprecedented example of how reintroduction of resistant/tolerant frogs obtained from recovering populations [@knapp2016] allows the successful reestablishment of extirpated MYL frog populations despite ongoing disease. Specifically, the reintroduction of resistant/tolerant frogs to 24 sites and results from long-term capture-mark-recapture (CMR) surveys indicate the capacity of reintroduced populations to become established and eventually recover. We used these data to parameterize a population model that extends our inferences well beyond the temporal extent of the CMR data, and shows that the recovering populations are likely to persist over a 50-year period. Finally, genomic results obtained using exome capture methods highlight evolutionary rescue as potentially key mechanism by which MYL frogs are able to recover in the presence of ongoing Bd infection. These results show that frogs have undergone substantial evolutionary change associated with Bd exposure, changes that might provide frogs with increased resistance/tolerance to disease. Collectively, these results provide one of the first examples of how evolutionary rescue might facilitate the successful reintroduction of resistant/tolerant animals and allow the large-scale recovery of disease-impacted species.-->

@@

<!--# I moved this paragraph to the top of results to provide an early focus on outcomes that are more obviously related to population recovery than survival. -->Of the `r nrow(d2)` translocated populations, `r frog_counts_gt0_nrow` (`r frog_counts_gt0_nrow / nrow(d2)`) showed evidence of successful reproduction in subsequent years, as indicated by the presence of tadpoles and/or juveniles.
For these `r frog_counts_gt0_nrow` populations, one or both life stages were detected in nearly all survey-years following translocation (proportion of survey-years: median = `r frog_counts_gt0_median`, range = `r frog_counts_gt0_min`--`r frog_counts_gt0_max`).
These same populations were also those in which recruitment of new adults (i.e., progeny of translocated individuals) was detected.
As with early life stages, recruits were detected in the majority of post-translocation survey-years (proportion of survey-years: median = `r frog_captures_gt0_median`, range = `r frog_captures_gt0_min`--`r frog_captures_gt0_max`).
In summary, survey results indicate that translocations resulted in the establishment of reproducing MYL frog populations at most recipient sites despite the ongoing presence of Bd.

@@

@@

Results from frog translocations indicated that most populations showed evidence of successful reproduction and recruitment and that adult survival was often relatively high (described above).
Although suggestive of population establishment, a decade or more of surveys may be necessary to confirm that populations are in fact self-sustaining [@joseph2018].
To extend our inferences of population establishment beyond those possible from the 2-16 years of site-specific CMR data, we developed a population viability model.
Specifically, to test whether the observed yearly adult survival probabilities in translocated populations were sufficient to promote long-term viability, we built a stage-structured matrix model that captured known frog demography and included demographic and environmental stochasticity.
We parameterized the model using CMR data from translocated populations and known life history values in this system (@tbl-param_values).

@@

Reintroduction of resistant/tolerant MYL frogs was remarkably successful in reestablishing viable populations in the presence of Bd.
Of the 12 translocated populations, approximately 80% showed evidence of both successful reproduction and recruitment of new adults.
Bd loads measured before and after translocation were similar and generally remained well below the level at which frog mortality occurs.
Annual survival for 12 of the 24 translocated cohorts exceeded 50%, and \> 70% of translocated cohorts had survival above this 50% level when the earliest translocations are excluded (i.e., translocations conducted when methods were still being refined).
The fact that the relatively low Bd loads and correspondingly high frog survival in the presence of Bd was maintained between the natal/donor sites and the recipient sites indicates that these characteristics of naturally-recovering populations were not solely an effect of the natal/donor site or site \* frog interactions, but instead were due at least in part to resistance/tolerance inherent in the frogs.
In addition, the relatively high survival of translocated frogs was maintained in their progeny, as expected if resistance/tolerance has a genetic basis.
Results from the population viability model were also encouraging.
In particular, translocated populations with \> 50% survival in the first year post-translocation were predicted to have a low probability of extinction over 50 years (probability of extinction = 0 - 0.32).
In light of the generally low success rate of amphibian reintroduction efforts [@dodd2005], our success in reestablishing MYL frog populations via translocation of resistant/tolerant individuals is striking, and even more so given that this success occurred despite ongoing infection of frogs with Bd.
<!--# Mention that 1-year survival can be used as an accurate proxy of long-term persistence.  -->
